Rough calculations

### 2016-01-30 drafts/hybrid_zone.tex
more analysis

\documentclass[12pt]{article}
\usepackage{geometry} % see geometry.pdf on how to lay out the page. There's lots.
\geometry{a4paper} % or letter or a5paper or ... etc
% \geometry{landscape} % rotated page geometry
\usepackage{natbib}
\usepackage{color}
% See the ``Article customise'' template for come common customisations
\newcommand{\alisa}[1]{{\em \color{red} #1}}

@@

@@

Consider a single locus, with alternate alleles fixed in the two parental populations,
for which heterozygotes have fitness $1-s$, and homozygotes are of equal fitness.
The deterministic theory predicts that after secondary contact,
these alleles will form a stable cline.
To study gene flow past this cline we first need to understand how the cline itself is formed.
Let $p(x,t)$ denote the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and that $p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
(In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
Let $\sigma^2$ be the mean squared distance between parent and offspring.
Assuming local random assortment of alleles into diploids,
that the habitat and dispersal are homogeneous,
and that $s$ and $\sigma$ are small, 
the commonly-used equation that $p$ solves is
\begin{align} \label{eqn:cline_pde}
    \dot p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\dot p(t,x)$ is the time derivative and $\Delta p$ is the Laplacian,
as derived in \citep{bazykin}; for discussion of diploidy see \citep{diploidcline}.
There is not an exact analytic solution to this equation, 
but a good approximation to the equilibrium solution is that
$\lim_{t \to \infty} p(x,t) \approx (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{bazykin}.
Considering the numerous approximations going into this,
the important attributes are that 
the stable cline has width of order $\sigma/\sqrt{s}$,
and decays exponentially.
Furthermore, the cline is established through diffusion.
This is slowed down somewhat by selection against heterozygotes,
but is still diffusive, and so it will take time of order $1/s$
to reach its stable width.
(To confirm this, 
rescale space by $\sigma/s$ and time by $1/s$ in equation \eqref{eqn:cline_pde},
and see that the result doesn't depend on $\sigma$ or on $s$.)

Now suppose that the frequency profile of the selected allele $p(x,t)$ is given.
We will understand how lineages of sampled individuals behave,
both at and away from the selected site,
by following lineages as they move between backgrounds,
as done for instance in \citet{selectioncoal,ralph2015patchy}.
Consider the collection of $A$ alleles found at location $x$.
The expected number of such alleles produced by location $y$ is proportional to 
the number of $A$ alleles at $y$ multiplied by the probability of dispersing from $y$ to $x$;
looking in the other direction, this is also the probability that an $A$ allele at $x$ had a parent at $y$.
In other words, lineages move as the random walk determined by dispersal,
but biased towards areas that produce more of the selected allele that they carry.
Making the same assumptions underlying equation \eqref{eqn:cline_pde},
we show in Appendix XXX
that the lineage of an $A$ allele moves as Brownian motion with speed $\sigma$
with drift $s \grad \log p(x,t)$ (i.e., it is pushed uphill on $\log p$ at speed $s$).
Since $p=1$ far from the zone to the left, 
and is proportional to $\exp(-x\sqrt{s}/\sigma)$ to the right,
then roughly, lineages on ``their own'' side wander randomly,
while lineages on ``the wrong'' side are pushed at constant speed $\sigma/\sqrt{s}$ 
back towards the side where they are more common.
Since an $A$ allele must by definition have been inherited from the $A$ side of the barrier 
at the time of secondary contact,
this push must get more intense the closer it is to the time of secondary contact.
Lineages of $B$ alleles do the same thing, of course, but in the opposite direction.

The behavior of a lineage at a linked locus is similar,
except that when in heterozygotes the lineage may move between backgrounds:
if we follow back through time the lineage of a locus we find linked to an $A$ allele, 
it will first tend to be inherited from ancestors to the left (as $A$ lineages drift to the left),
but if it spends time in the clinal region where recombination may happen,
it may be found to have been inherited from a $B$-carrying individual,
whose ancestors will tend to be more from the right.
Suppose we sample an allele today at a locus at recombination distance $r$ to the selected site.
If $X_t$ is the location of its ancestor $t$ generations ago,
and $Z_t$ is the identity of the selected allele that ancestor carried,
then we say that $X$ moves as a diffusion pushed by either $\log(p)$ or $\log(1-p)$,
and that $Z$ jumps between $A$ and $B$ at rate either $r p$ or $r(1-p)$,
depending on the identity of $Z$.
In symbols,
\begin{align}
    \begin{aligned}
        dX_t &= \begin{cases}
            & s \grad \log(p(X_t,t)) dt + dB_t \qquad & \text{if } Z_t = A \\
            & s \grad \log(1-p(X_t,t)) dt + dB_t \qquad & \text{if } Z_t = B 
        \end{cases} \\
        \P\{ Z_{t+\epsilon} &= B \given Z_t = A \} = \epsilon r (1-p(X_t,t)) + O(\epsilon^2) \\
        \P\{ Z_{t+\epsilon} &= A \given Z_t = B \} = \epsilon r p(X_t,t) + O(\epsilon^2)  .
    \end{aligned}
\end{align}

\textbf{illustrative figure of a lineage moving back and forth?}

Clines at linked sites are given by
the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
Denote this probability $q_r(x,t,z)$.
In other words,
\begin{align}
    q_r(x,t,z) = \P^x \{Z_t = A\} .
\end{align}
The description above implies that $q_r$ solves the equation
\begin{align}
    \dot q_r(x,t,A) &= \grad \log(p(x,t)) \cdot \grad q_r(x,t,A) 
            + \frac{\sigma^2}{2} \Delta q_r(x,t,A)
            + r (1-p(x,t))(q_r(x,t,B)-q_r(x,t,A)) \\
    \dot q_r(x,t,B) &= \grad \log(1-p(x,t)) \cdot \grad q_r(x,t,B) 
            + \frac{\sigma^2}{2} \Delta q_r(x,t,B)
            + r (1-p(x,t))(q_r(x,t,A)-q_r(x,t,B))  ,
\end{align}
with boundary conditions $q_r(x,0,A)=1$ and $q_r(x,0,B)=0$.

@@

@@

>>>>

\documentclass[12pt]{article}
\usepackage{geometry} % see geometry.pdf on how to lay out the page. There's lots.
\geometry{a4paper} % or letter or a5paper or ... etc
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{natbib}
\usepackage{color}

\newcommand{\alisa}[1]{{\em \color{red} #1}}
\newcommand{\plr}[1]{{\em \color{blue} #1}}

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\one}{\mathbf{1}}
\DeclareMathOperator{\sgn}{sgn}
\newcommand{\grad}{\nabla}

\newcommand{\given}{\,\vert\,}
\newcommand{\st}{\,\colon\,}
\renewcommand{\and}{\,\&\,}

@@

The tracking lineages given selection trick is an example of the more general idea of duality for particle systems.
It also assumes a large pop density limit that excludes recombination and makes lineages Markov;
if things are very patchy this might not be a good one.
Also, parent-offspring dispersal is not the same thing as successful-offspring-parent dispersal.

@@

\plr{The aim here is to give both the heuristics, for how a lineage moves around
and what that tells us about the scale on which clines happen;
as well as writing down the PDE.  These should probably be better separated?}

Perhaps start out with the main points:
the cline takes time $s$ to spread out,
and after time $\tau$ will have clines out to genetic distance of order $1/\tau$,
maybe more.

\paragraph{Establishment of the selected cline}
Consider a single locus, with alternate alleles fixed in the two parental populations,
for which heterozygotes have fitness $1-s$, and homozygotes are of equal fitness.
The deterministic theory predicts that after secondary contact,
these alleles will form a stable cline.
To study gene flow past this cline we first need to understand how the cline itself is formed.
Let $p(x,t)$ denote the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and that $p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
(In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
Let $\sigma^2$ be the mean squared distance between parent and offspring.
Assuming local random assortment of alleles into diploids,
that the habitat and dispersal are homogeneous,
and that $s$ and $\sigma$ are small, 
the commonly-used equation that $p$ solves is
\begin{align} \label{eqn:cline_pde}
    \dot p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\dot p(t,x)$ is the time derivative and $\Delta p$ is the Laplacian,
as derived in \citep{bazykin}; for discussion of diploidy see \citep{diploidcline}.
There is not an exact analytic solution to this equation, 
but a good approximation to the equilibrium solution is that
$\lim_{t \to \infty} p(x,t) \approx (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{bazykin}.
Considering the numerous approximations going into this,
the important attributes are that 
the stable cline has width of order $\sigma/\sqrt{s}$,
and decays exponentially.
Furthermore, the cline is established through diffusion.
This is slowed down somewhat by selection against heterozygotes,
but is still diffusive, and so it will take time of order $1/s$
to reach its stable width.
(To confirm this, 
rescale space by $\sigma/s$ and time by $1/s$ in equation \eqref{eqn:cline_pde},
and see that the result doesn't depend on $\sigma$ or on $s$.)

\paragraph{Lineages at the selected locus}
Now suppose that the frequency profile of the selected allele $p(x,t)$ is given.
We will understand how lineages of sampled individuals behave,
both at and away from the selected site,
by following lineages as they move between backgrounds,
as done for instance in \citet{selectioncoal,ralph2015patchy}.
Consider the collection of $A$ alleles found at location $x$.
The expected number of such alleles produced by location $y$ is proportional to 
the number of $A$ alleles at $y$ multiplied by the probability of dispersing from $y$ to $x$;
looking in the other direction, this is also the probability that an $A$ allele at $x$ had a parent at $y$.
In other words, lineages move as the random walk determined by dispersal,
but biased towards areas that produce more of the selected allele that they carry.
Making the same assumptions underlying equation \eqref{eqn:cline_pde},
we show in Appendix \ref{apx:lineage_derivation}
that the lineage of an $A$ allele moves as Brownian motion with speed $\sigma$
with drift $s \grad \log p(x,t)$ (i.e., it is pushed uphill on $\log p$ at speed $s$).
Since $p=1$ far from the zone to the left, 
and is proportional to $\exp(-x\sqrt{s}/\sigma)$ to the right,
then roughly, lineages on ``their own'' side wander randomly,
while lineages on ``the wrong'' side are pushed at constant speed $\sigma/\sqrt{s}$ 
back towards the side where they are more common.
Since an $A$ allele must by definition have been inherited from the $A$ side of the barrier 
at the time of secondary contact,
this push must get more intense the closer it is to the time of secondary contact.
Lineages of $B$ alleles do the same thing, of course, but in the opposite direction.

\paragraph{Lineages at linked loci}
The behavior of a lineage at a linked locus is similar,
except that when in heterozygotes the lineage may move between backgrounds:
if we follow back through time the lineage of a locus we find linked to an $A$ allele, 
it will first tend to be inherited from ancestors to the left (as $A$ lineages drift to the left),
but if it spends time in the clinal region where recombination may happen,
it may be found to have been inherited from a $B$-carrying individual,
whose ancestors will tend to be more from the right.
Suppose we sample an allele today at a locus at recombination distance $r$ to the selected site.
If $X_t$ is the location of its ancestor $t$ generations ago,
and $Z_t$ is the identity of the selected allele that ancestor carried,
then we say that $X$ moves as a diffusion pushed by either $\log(p)$ or $\log(1-p)$,
and that $Z$ jumps between $A$ and $B$ at rate either $r p$ or $r(1-p)$,
depending on the identity of $Z$.
In symbols,
\begin{align}
    \begin{aligned} \label{eqn:lineage_motion}
        dX_t &= \begin{cases}
             s \grad \log(p(X_t,t)) dt + dB_t \qquad & \text{if } Z_t = A \\
             s \grad \log(1-p(X_t,t)) dt + dB_t \qquad & \text{if } Z_t = B 
        \end{cases} \\
        \P\{ Z_{t+\epsilon} = B \given Z_t = A \} &= \epsilon r (1-p(X_t,t)) + O(\epsilon^2) \\
        \P\{ Z_{t+\epsilon} = A \given Z_t = B \} &= \epsilon r p(X_t,t) + O(\epsilon^2)  .
    \end{aligned}
\end{align}

\textbf{figure of a lineage moving back and forth?}

\paragraph{How fast do linked clines relax?}
Although loci not under selection can in principle spread across the cline unimpeded,
in practice it can take quite some time for even unlinked neutral loci to homogenize,
due to the decreased fitness of heterozygotes \citep{bartonbarrier}
and the relative slowness of diffusive movement \citep{sedghifar}.
Suppose we sample across at cline $\tau$ units of time after secondary contact.
How far away from the selected locus do we expect to see clines that have begun to flatten out more
than the selected cline?
A linked cline will have begun to flatten out if there is an appreciable chance that
a lineage at the linked locus initially linked to an $A$ allele 
at some point has recombined from an individual carrying the $B$ allele.
This implies that loci closer than $1/\tau$ to the selected site
certainly won't have a substantially wider cline than the selected cline.
\plr{Figure showing this?}
Of course, this calculation assumes that every generation provides an opportunity for recombination 
onto the other background,
but only recombinations in heterozygotes will move the linked lineage between selected backgrounds.
Since heterozygotes for the selected allele are only found at high frequency in the cline,
and lineages are generally pushed away from the cline but have no bias far away,
the amount of time a lineage spends in heterozygotes should grow as $\sqrt{\tau}$ for large $\tau$,
and so the width of the genomic region showing clines about the selected locus could be 
substantially larger than $1/\tau$.
However, this distinction appears hard to observe for realistic parameter values.

\paragraph{PDE for linked clines}
Clines at linked sites are given by
the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
Denote this probability $q_z(x,t,r)$.
In other words,
\begin{align}
    q_z(x,t,r) = \P^x \{Z_t = A\} .
\end{align}
The description above implies that $q_r$ solves the equation
\begin{align}
    \begin{aligned}  \label{eqn:q_pde}
    \dot q_A(x,t,r) 
            &= \grad \log(p(x,t)) \cdot \grad q_A(x,t,r) 
                + \frac{\sigma^2}{2} \Delta q_A(x,t,r) 
            \\ &\qquad {} + 
                r (1-p(x,t))(q_B(x,t,r)-q_A(x,t,r))  \\
    \dot q_B(x,t,r) &= \grad \log(1-p(x,t)) \cdot \grad q_B(x,t,r) 
            + \frac{\sigma^2}{2} \Delta q_B(x,t,r)
            \\ &\qquad {} + 
            + r (1-p(x,t))(q_A(x,t,r)-q_B(x,t,r))  ,
    \end{aligned} \\
\end{align}
with boundary conditions $q_A(x,0,r)=1$ and $q_B(x,0,r)=0$.
We will have more use for the differential operators on the right-hand sides of these equations,
so write these also as 
$\dot q_A = G_r^A q_A + r (1-p) (q_A-q_B)$
and 
$\dot q_B = G_r^B q_B + r (1-p) (q_B-q_A)$.

\paragraph{Numerical computation}
To determine how the cline at a linked locus is expected to relax,
we can solve the partial differential equations \eqref{eqn:cline_pde} and \eqref{eqn:q_pde} numerically,
producing, for instance, a heatmap of expected local frequencies of $A$ ancestry
across space and time, as in Figure \ref{fig:pde_clines}.
Shown is the expected frequency of ancestry $A$ at location $x$ and time $t$ at a site at recombination distance $r$ from the selected site,
which we denote $p(x,t,r)$ and can compute as $p(x,t,r) = p(x,t) q_A(x,t,r) + (1-p(x,t)) q_B(x,t,r)$.

%%%%%%
\begin{figure}
    \begin{center}
        \includegraphics[width=0.7\textwidth]{figs/example_cline}
    \end{center}
    \caption{
        \textbf{Numerically computed clines,} at
        \textbf{(A)} the selected site, which forms over time $1/s$ and then remains stable, and
        \textbf{(B)} a linked site at distance $r=0.1$, which continues to smooth out as $\sqrt{t}$.
        Other parameters.
        \plr{Do we want this figure? This one is just a placeholder; I'd put in two, that showed the formation and spread better.
        Maybe putting in simulation results would be better, with comparison to numerical solutions? But, this is easier to look at?}
        \label{fig:pde_clines}
    }
\end{figure}

\paragraph{Haplotype lengths}
We can also find the probability of seeing an entire block of genome (haplotype)
of a single ancestry at a given location and time.
As in \citet{sedghifar},
to do this we need only keep track of both lineages when a recombination in the middle of a block occurs,
so that lineages will act like a particular class of labeled, branching diffusions,
where the total branching rate is conserved.
We will not consider subsequent coalescence.
The general description of the process is as follows.
The lineage of a segment of genome behaves as a linked locus described in equations \eqref{eqn:lineage_motion},
with recombination distance $r$ equal to the rate at which the segment recombines away from the selected site.
(If the selected site lies inside the segment, $r=0$.)
Additionally, at rate equal to the genetic length of the segment,
a recombination occurs, at which point the segment splits into two segments at a uniformly chosen location,
each of which procedes as before, independently.

\paragraph{Spatial haplotype patterns}
Furthermore, we expect that stretches of $A$ ancestry will tend to be shorter the further one moves into the $B$ side of the cline,
because they have had more opportunities to recombine with $B$ haplotypes.
However, stretches of $A$ ancestry that contain the selected site are expected to be longer
than those that do not at the same spatial location,
because lineages containing the selected site must have been inherited from the $A$ side of the cline recently --
concretely, they move at speed roughly $\sigma/\sqrt{s}$,
so we expect (selected) $A$ alleles at distance $x$ from the cline center
to have last had an ancestor on the $A$ side of the cline around $x \sqrt{s}/\sigma$ generations ago,
and so the enclosing $A$ haplotype is no longer than $\sigma/(x \sqrt{s})$.
\plr{(check predicted decay of mean length $1/x$ in simulations)}
For linked haplotypes not containing the selected site, the situation is more complicated, and harder to intuit.

\paragraph{General haplotype probabilities}
To quantify haplotype abundances,
we'd like to know the probability that the entire segment of genome between positions $a$ and $b$ relative to the selected site
in an individual sampled at spatial location $x$ and time $t$ after the cline has formed
is of ancestry $A$.
Writing this probability $g_A(x,t;a,b)$,
branching adds a term to equation \eqref{eqn:q_pde}, so that 
\plr{write this equation more compactly?}
\begin{align}
    \begin{aligned} \label{eqn:g_pde}
        \dot g_A(x,t;a,b) 
            &= G_{r(a,b)}^A g_A(x,t;a,b) 
            - (b-a) g_A(x,t;a,b) 
            \\ {} & \qquad 
            + p(x,t) \int_a^b {
                g_A(x,t;a,\theta) g_A(x,t;\theta,b) 
                - g_A(x,t;a,b) .
            } d\theta
            \\ {} & \qquad 
            + (1-p(x,t)) \int_a^0 {
                g_B(x,t;a,\theta) g_A(x,t;\theta,b)
            } d\theta
            \\ {} & \qquad 
            + (1-p(x,t)) \int_0^b {
                g_A(x,t;a,\theta) g_B(x,t;\theta,b)
            } d\theta,
    \end{aligned}
\end{align}
where the integral $\int_a^0$ is zero if $a>0$ and the integral $\int_0^b$ is zero if $b<0$;
and $r(a,b)$ is the distance from the segment to the selected site:
$r(a,b)=0$ if $ab<0$ and $r(a,b)=\min(|a|,|b|)$ otherwise.
The equation for $g_B(x,t;a,b)$ is identical after exchanging $A \leftrightarrow B$ and
$p \leftrightarrow (1-p)$.
The boundary conditions are $g_A(x,0;a,b)=1$ and $g_B(x,0;a,b)=0$.

\paragraph{Numerical solutions}
Notice that equations \eqref{eqn:g_pde}
are hierarchical in $[a,b]$:
the equation for haplotype identity probabilities on a segment $[a,b]$ depends only on those probabilities for smaller segments.
This allows for efficient numerical solutions,
described in more detail in Appendix \ref{apx:haplotype_calcs}.

%%%%%%
\begin{figure}
    \begin{center}
    \end{center}
    \caption{
        \textbf{Haplotype length distributions},
        comparing numerical solutions to simulations.
        \plr{maybe?}
        \label{fig:haplotype_lengths}
    }
\end{figure}

\paragraph{Correlations in ancestry}
To compute correlations in local ancestry
(i.e., ``ancestry disequilibrium'', as in \citet{pool,brandvainthisvolume}),
we need only follow lineages at two sites, instead of an entire region,
the model described above requires only minor modifications,
and our numerical code can be used directly.

@@

Numerical solution for the PDE can be used on real, non-flat landscapes that change with time;
we have code and notes on solving them on github.

@@

\appendix

\section{Rescaling a dicrete model to obtain the lineage motion}
\label{apx:lineage_derivation}

Here put derivation of the PDE for $p(x,t)$ and also the SDE of a lineage motion.

\section{Nuemrical solutions for haplotype probabilities}
\label{apx:haplotype_calcs}.

Here put some details about the numerical solutions of haplotype probabilities.

### 2016-01-30 notes/rough-calculations.Rmd
more analysis

d.  By Lévy's identity, the local time of a Brownian motion at zero
    has the same distribution (pathwise!) as the running maximum of Brownian motion.
    Therefore, the local time at zero up until $\tau$ grows like $\sqrt{\tau}$.
    The same is true for Brownian motion reflected at zero.
    The time a lineage spends in heterozygotes should be roughly proportional to the local time
    (each time it visits zero from the left, it spends some time in hets before leaving again; 
    the constant of proportionality is the mean time spent per visit).
    Therefore, there will be appreciable chance of recombination if $r C \sqrt{\tau}>1$,
    where $C$ is the constant;
    this says that if clines down to $r$ are flat at $\tau$, then at time $2\tau$ they'll be flat down to $r/4$.

>>>>

d.  By Lévy's identity, the local time of a Brownian motion at zero
    has the same distribution (pathwise!) as the running maximum of Brownian motion.
    Therefore, the local time at zero up until $\tau$ grows like $\sqrt{\tau}$.
    The same is true for Brownian motion reflected at zero.
    The time a lineage spends in heterozygotes should be roughly proportional to the local time
    (each time it visits zero from the left, it spends some time in hets before leaving again; 
    the constant of proportionality is the mean time spent per visit).
    Therefore, there will be appreciable chance of recombination if $r C \sqrt{\tau}>1$,
    where $C$ is the constant;
    this says that if clines down to $r$ are flat at $\tau$, then at time $4\tau$ they'll be flat down to $r/2$.
